NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PS lipids

### 2018-01-24 Manuscript/manuscriptPS.tex
Update manuscript

%\label{}
\section{Introduction}
Cellular membranes are composed of mixtures of zwitterionic
and negatively charged lipids. Negatively charged lipids are important for
the signalling and lateral organization of cellular membranes \cite{li4}.
Direct interactions of cations and proteins with negatively charged
lipids have been proposed to trigger signaling events \cite{li14,??}.
On the other hand, it has been proposed that negatively charged lipids 
only increase the local concentration of cations in the vicinity of
the membrane and that the binding constant of cations is similar
to zwitterionic and negatively charged lipids \cite{seelig90,??}.

Details of negatively charged lipids and their interactions
with ions and proteins have been experimentally analyzed with
various spectroscopic methods \cite{borle85,macdonald87,roux90,seelig90,pan14,melcrova16,??}
%NMR \cite{seelig90,??}, scattering \cite{pan14,??} and fluorescence spectroscopy
which are often complemented with classical MD simulations \cite{pan14,melcrova16,??}.
The structure of negatively charged lipids and their detailed
interactions with ions or proteins have not, however, been fully resolved
due to the difficulties to interpret the experimental data \cite{??} and
the inaccuracies in simulation models at the
headgroup region \cite{botan15,catte16,ollila16}.

Headgroup and glycerol backbone C-H bond order parameters
calculated from MD simulations have been
recently used to interpret the lipid structures in NMR experiments
and to validate lipid structure and ion binding in simulations of
PC lipid bilayers \cite{botan15,catte16,ollila16,ferreira16}.
In this work we apply the same approach to negatively charged
phosphatidylserine lipids (PS) which compose   
25-35\% of plasma membrane \cite{li14,??}, regulate
surface charge and protein localization \cite{yeung08}, and
induce protein aggregation \cite{zhao04,??}.

@@

Glycerol backbone and headgroup structure and behaviour are similar
in model membranes and in bacteria \cite{gally81,scherer87,seelig90}.

@@

\todo{When we have more data for Ca binding to PS containing bilayers, the discussion will be updated and PG results moved
to other manuscript.}
Comparison of Ca$2+$ binding in PG between CHARMM36 simulations and experiments \cite{borle85}
is shown in Fig. \ref{changesWITHCaClPG}. The decrease of $\alpha$ order parameter
is in agreement with experiments, while decerase of $\beta$ order parameter is overestimated.
The result is very similar to the results with PC in NMRlipids II publication \cite{catte16}.
It should be, however, noted that the $\beta$-order parameters are not actually measured for PG,
but they are calculated from empirical relation $\Delta S_{\beta}=0.43\Delta S_{\alpha}$ \cite{akutsu81}.
Anyway, the data presented in NMRlipids II project and in Fig. \ref{changesWITHCaClPG} together
suggest that Calcium binding is similarly overestimated by CHARMM36 model in pure POPC bilayers and mixtures with
POPG. The good agreement of $\alpha$ carbon would be explained by too weak dependence of its order
parameter of bound charge
\begin{figure*}[]
  \centering
  \includegraphics[width=18cm]{../Figs/CHANGESwithCaClPS.eps}
  \caption{\label{changesWITHCaClPG}
    PG order parameters as a function CaCl$_2$ concentration from experiments \cite{borle85} and CHARMM36 simulations.
    Note that beta order parameter is calculated from empirical relation $\Delta S_{\beta}=0.43\Delta S_{\alpha}$ \cite{akutsu81}, not actually measured. 
  }
\end{figure*}

@@

\todo{More simulation data for systems with negatively charged lipids and CaCl$_2$ to be collected}

@@

Also the experimental order parameters for PS and PG headgroups
as a function of CaCl$_2$ concentration are shown in Fig. \ref{PSPGchangesWITHCaCl}.
\todo{These should be compared to simulations for potential structural interpretation of the changes.}
%\begin{figure}[]
%  \centering
%  \includegraphics[width=9.0cm]{../Figs/PSPGwithCaCl.eps}
%  \caption{\label{PSPGchangesWITHCaCl}
%    PG and PS order parameters a function CaCl$_2$ concentration taken from \cite{borle85} and \cite{roux90}, respectively.
%  }
%  \todo{Get the small concentration data from the inserts}
%\end{figure}

@@

% If you have acknowledgments, this puts in the proper section head.
\begin{acknowledgments}
% Put your acknowledgments here.
\end{acknowledgments}
\newpage
\appendix
\begin{center}
{\bf SUPPLEMENTARY INFORMATION}
\end{center}

>>>>

%\label{}
\section{Introduction}
Phosphatidylserine (PS) is the most common negatively
charged lipid in eykaryotic membranes.
PS lipids compose 8.5\% of total lipid weight of erythrocytes,
but the abundance varies between different organelles up to
25-35\% in plasma membrane \cite{lemmon08,leventis10,li14}.
Despite of the relatively low abundance, PS lipids
are important signaling molecules. They interact with
signaling proteins \cite{leventis10}, regulate
surface charge and protein localization \cite{yeung08}, and
induce protein aggregation \cite{zhao04,??}.
Some domains spesifically interact PS lipids,
while others are attracted by general electrostatics and the
binding can be regulated by calcium \cite{leventis10}.
Therefore, it is crucial to resolve the structural details
of PS headgroup and calcium binding in order to fully
understand PS mediated signaling processes.

The structure of PS lipid headgroups and their 
interactions with ions have been studied with
various experimental methods and theoretical techniques \cite{roux90,melcrova16,??}.
However, the consesus has not been reached 
due to the difficulties to interpret the experimental data \cite{??} and
the inaccuracies in simulation models at the
headgroup region \cite{botan15,catte16,ollila16}.
Some studies propose specific binding of calcium directly to PS lipid
headgroups \cite{??}, while others propose that only the local concentration
of cations in the vicinity of negatively charged membranes containing PS lipids
and that the binding constant of cations is similar
to zwitterionic and negatively charged lipids \cite{seelig90,??}.
The NMR data proposes that the PS headgroup is more rigid than PC,PE or PG
headgroups, but more detailed interperetation has not been done.

%Details of negatively charged lipids and their interactions
%with ions and proteins have been experimentally analyzed with
%various spectroscopic methods \cite{borle85,macdonald87,roux90,seelig90,pan14,melcrova16,??}
%NMR \cite{seelig90,??}, scattering \cite{pan14,??} and fluorescence spectroscopy
%which are often complemented with classical MD simulations \cite{pan14,melcrova16,??}.

Headgroup and glycerol backbone C-H bond order parameters
calculated from MD simulations have been
recently used to interpret the lipid structures in NMR experiments
and to validate lipid structure and ion binding in simulations of
PC lipid bilayers \cite{botan15,catte16,ollila16,ferreira16}.
In this work we apply this approach to PS lipid headgroup
in order to elucidate the structural details and ion binding
to negatively charged lipids. The results are expected to elucidate
also PS mediated signalling events because 
glycerol backbone and headgroup structure and behaviour are similar
in model membranes and in bacteria \cite{gally81,scherer87,seelig90}.

@@

@@

The headgroup order parameter changes of POPC and POPS from POPC:POPS (5:1) bilayers
as a function of Ca$2+$ concentration from different
simulations and experiments \cite{roux90} are shown in Fig. \ref{changesWITHCaClPG}.
As shown previously, the changes of PC headgroup order parameters can be
used to measure the amount of bound ions to the bilayer according to the
electrometer concept \cite{akutsu81,seelig87,catte16}.
This would suggest that the Ca2+ binding is overestimated in
MacRog simulation and underestimated in CHARMM36 simulation.
It should be however noted, that MacRog simulation with low
concetrations gives a good agreemet with experiments.
The result from CHARMM simulation is somewhat surprising because
previous study suggested that Ca2+ ions overbind to pure PC lipid
bilayers \cite{catte16}.
\todo{The role of counterions and equilibration should be somehow checked
and discussed.}
\begin{figure*}[]
  \centering
  \includegraphics[width=18cm]{../Figs/CHANGESwithCaClPS.eps}
  \caption{\label{changesWITHCaClPG}
    POPC (left) and POPS (right) headgroup order parameters in lipid bilayer with POPC:POPS (5:1) mixture
    as a function CaCl$_2$ concentration from experiments \cite{roux90} and different simulations.
  }
\end{figure*}

@@

\todo{More simulation data for systems with PS lipids and CaCl$_2$ would be probably useful}

@@

Also the experimental order parameters of POPS headgroup 
as a function of CaCl$_2$ concentration are shown in Fig. \ref{PSPGchangesWITHCaCl}.
The order parameter changes with added CaCl$_2$ are overestimated in
both simulation models. However, the changes can be considered qualitatively
correct for both segments in CHARMM36 model and in $\beta$ segment in MacRog model. 
\todo{These should be compared to simulations for potential structural interpretation of the changes.}
%\begin{figure}[]
%  \centering
%  \includegraphics[width=9.0cm]{../Figs/PSPGwithCaCl.eps}
%  \caption{\label{PSPGchangesWITHCaCl}
%    PG and PS order parameters a function CaCl$_2$ concentration taken from \cite{borle85} and \cite{roux90}, respectively.
%  }
%  \todo{Get the small concentration data from the inserts}
%\end{figure}

@@

% If you have acknowledgments, this puts in the proper section head.
\begin{acknowledgments}
% Put your acknowledgments here.
\end{acknowledgments}
\pagebreak
\appendix
\begin{center}
{\bf SUPPLEMENTARY INFORMATION}
\end{center}
